---
title: "EDLD 651 Final Project"
author: "Jon Benedik Bunquin, Nafisul Huq, and Faria Shaikh"
format: html
editor: visual
warning: False
embed-resources: true
toc: true
bibliography: references.bib
---

# Abstract

# Introduction

**Introduction**

The geopolitical landscape of the post-World War II era witnessed a monumental shift with the global wave of decolonization, especially pronounced in the 1960s. This period marked the birth of numerous newly sovereign states, many of which embarked on the path of democratic governance. These nascent democracies, emerging from the shadows of
colonial rule, faced the daunting task of establishing democratic systems, cultivating a political culture rooted in democratic principles, and nurturing values conducive to a democratic ethos. However, as illuminated by Kapstein and Converse (2008), these democracies, in their formative stages, encountered
significant challenges, including institutional fragility and vulnerability to internal and external pressures, leading to concerns about democratic backsliding -- a regressive shift from democratic norms that could potentially pave the way for authoritarian regimes.

**Review of Literature**

The study of electoral democracy over the past fifty years, especially in the context of post-colonial countries and comparative analyses with established democracies like the United States,
is a rich field of inquiry. This literature review will explore key themes including the global context of democratic evolution post-decolonization, the specific democratic trajectories of Bangladesh, Pakistan, and the Philippines, methodological approaches in democracy research, and predictive models in
political science.

**Decolonization and Democracy**

Smith (2010) offers an overview of political transformations following decolonization, emphasizing the complexities faced by new states in establishing democratic governance. His
work sheds light on the interplay between ethnic diversity, economic
challenges, and the legacies of colonial rule. This perspective aligns with the multifaceted challenges and transitions experienced by post-colonial societies.
(1991) provides a seminal perspective on global democratization trends, including the challenges and successes of new democracies during this period.
While not directly focused on decolonization, 's work contributes to understanding the broader context of political transformations and democratization in post-colonial states, which is relevant to the complexities of decolonization
processes. "Decolonization: a short history" offers insights into the
historical context and processes of decolonization, providing a foundational understanding of the subject. This historical perspective is valuable for comprehending the evolution of decolonization and its impact on political
science and colonization.

Datta (2017) discusses the effectiveness
of decolonizing both researcher and research in Indigenous research, highlighting the empowerment that comes with decolonization. This perspective is crucial for understanding the transformative potential of decolonization in
research and academia, particularly in the context of Indigenous knowledge systems. Sultana (2019) explores the pursuit of social justice through decolonizing development education, emphasizing the liberation and new ways of valuing that are central to decolonization. This perspective contributes to understanding the broader implications of decolonization beyond political
contexts, encompassing ideologies and praxis. Spiegel et al. (2016) reflect on efforts to 'decolonize development studies' and the complex power dynamics they encounter, providing insights into the challenges and aspirations of decolonizing scholarship. This critical reflection is valuable for understanding the complexities and power dynamics involved in decolonization efforts within
academic disciplines.

**Democratic Backsliding**

The phenomenon of democratic backsliding has garnered significant scholarly attention, with a focus on understanding the erosion of democratic norms and institutions Bermeo (2016). discusses democratic backsliding, emphasizing institutional vulnerabilities and
the erosion of democratic norms and structures. This work sheds light on the challenges faced by democracies in maintaining their democratic character and the factors contributing to democratic erosion. 
Riaz (2016) delves into Bangladesh's political and democratic journey, focusing on its struggles with military rule and democratic governance. This country-specific analysis provides insights into the unique challenges and trajectories of Bangladesh's democracy, contributing to the understanding of
democratic backsliding in the context of Bangladesh.  Whereas Jalal (2014) offers a critical analysis of Pakistan's democratic evolution, highlighting the interplay between civilian governance and military power.
Hutchcroft (2016) examines the role of political dynasties and the
influence of populist leadership in shaping the Philippines' democratic framework.

**Comparative Analysis of Democracies**

Lijphart's work "Patterns of Democracy" provides a comprehensive framework for understanding different models of democracies and their comparative performances. This seminal work has
been influential in shaping the study of democratic systems and their variations, contributing to a deeper understanding of the complexities of
democratic governance.  Williams (2018)
discusses the application of ARIMA models in political forecasting, offering an approach that can be instrumental in predicting future trends in electoral democracies.

# Methods

```{r}
## Imports

#| echo: false

#install.packages("devtools")
#install.packages("vdemdata")

library(vdemdata)
library(dplyr)
library(skimr)
library(tidyverse)
library(ggplot2)
library(stats)
vdem <- vdem
```

## Data Processing

### Data Filtering

```{r}
vdemdata <- vdem %>%
  filter(country_name == "Bangladesh" | 
           country_name == "Philippines" |
           country_name == "Pakistan" |
           country_name == "United States of America") %>%
  select(country_name, 
         year, 
         v2x_frassoc_thick,
         v2xel_frefair,
         v2x_freexp_altinf,
         v2x_elecoff,
         v2x_suffr) %>%
  filter(year >= 1971) %>% 
  rename(freedom_association=v2x_frassoc_thick,
         clean_elections=v2xel_frefair,
         freedom_expression=v2x_freexp_altinf,
         elected_index=v2x_elecoff,
         suffrage=v2x_suffr)

head(vdemdata)

```

### Data Exploration

```{r}
skim(vdemdata)
```

<insert interpretation here>

### Data Cleaning

```{r}
vdemdata_long <- vdemdata %>%
  pivot_longer(
    cols = 3:5,
    names_to = 'criteria',
    values_to = 'score'
    )

head(vdemdata_long)
```

# Results

## Individual Country Results

### Bangladesh

```{r}
vdemdata_long %>%
  filter(country_name == 'Bangladesh') %>% 
  ggplot(aes(x = year, y = score, color = criteria)) +
  geom_point() +
  geom_line() + 
  geom_vline(xintercept = c(1975,1980,1990,2000,2007,2013)) + 
  annotate("text", x=c(1974,1979,1989,1999,2006,2012), y=0.1,label=c("Before 1975","1980","1990","2000","2007","2013"), 
           color="tomato4",size=2.5, angle=90) + 
  labs(title = 'Electoral Democracy in the Bangladesh from 1971 to 2022',
       subtitle = 'Trends in electoral democracy index scores over the last 50 years',
       x = 'Year',
       y = 'Index Score',
       color = 'Index',
       caption = 'Data Source: Varieties of Democracy (https://www.v-dem.net)') +
  scale_color_manual(values = c("#d8b365", "#2c7fb8", "#5ab4ac"),
                    labels = c('Clean election', 'Freedom of association', 'Freedom of expression')) + 
  theme_minimal() +
  theme(plot.caption = element_text(hjust = 0))
  
```

The above graph of Bangladesh shows an overview of the country's electoral democracy status from **1970 to 2022** focusing on three key variables - freedom of association, freedom of expression, and clean election. It reveals an unwavering trend of electoral democracy in the country ultimately leaning towards a declining state with a cumulative effect of the variables, where freedom of association is seen in a consistent battle to stay resilient in its consistent unstable trajectory, freedom of expression is found experiencing the most fluctuations with more decline and less rise in scores throughout the interval, and clean election is seen affected by the most severe slumps.\
</br> The ups and downs of the variables include many significant events that impacted the democracy in Bangladesh in the given period. To begin with, Bangladesh witnessed the first blow in **1975** with the formation of a one-party government that threatened the country's free and fair election, freedom of expression, and freedom of association. Then, it saw the emergence of military rule in late **1975** against the backdrop of tyrannical rule lasting through **1980**. Although sounds weird, it appeared as a pathway to get back to democracy. After years of smooth functioning, the military rule transformed into a figure of hatred and disdain. As a result, Bangladesh saw the fall of military dictatorship and the rise of democracy from **1990 to 2000**. The next bar towards the country's democracy became apparent in **2007** when an interim government came into play to offset that bar. While democracy sustained after that period it did not sustain for long. Greed of the erstwhile ruling party to stay in power battered the foundations of democracy, including clean election, freedom of speech, and freedom of association along with other components. The consequence of this event was so adverse that the country kept struggling to ascend the ladder of democracy until recently.

### Pakistan

```{r}

vdemdata_long %>%
  filter(country_name == 'Pakistan') %>% 
  ggplot(aes(x = year, y = score, color = criteria)) +
  geom_point() +
  geom_line() + geom_vline(xintercept = c(1971,1977,1988,1999,2008)) + 
  annotate("text", x=c(1970,1976,1986,1998,2007), y=0.1,label=c("Before 1971","1977","1988","1999","2008"),
           color="tomato4",size=2.5, angle=90) +
  labs(title = 'Electoral Democracy in the Pakistan from 1971 to 2022',
       subtitle = 'Trends in electoral democracy index scores over the last 50 years',
       x = 'Year',
       y = 'Index Score',
       color = 'Index',
       caption = 'Data Source: Varieties of Democracy (https://www.v-dem.net)') +
  scale_color_manual(values = c("#d8b365", "#2c7fb8", "#5ab4ac"),
                    labels = c('Clean election', 'Freedom of association', 'Freedom of expression')) + 
  theme_minimal() +
  theme(plot.caption = element_text(hjust = 0))

```

### Philippines

```{r}
vdemdata_long %>%
  filter(country_name == 'Philippines') %>% 
  ggplot(aes(x = year, y = score, color = criteria)) +
  geom_point() +
  
  geom_line() + 
  geom_vline(xintercept = c(1970,1986,2004,2009,2016)) + 
  annotate("text", x=c(1969,1985,2003,2008,2015), y=0.1,label=c("1970","1986","2004","2009","2016"), 
           color="tomato4",size=2.5, angle=90) +
  labs(title = 'Electoral Democracy in the Philippines from 1971 to 2022',
       subtitle = 'Trends in electoral democracy index scores over the last 50 years',
       x = 'Year',
       y = 'Index Score',
       color = 'Index',
       caption = 'Data Source: Varieties of Democracy (https://www.v-dem.net)') +
  scale_color_manual(values = c("#d8b365", "#2c7fb8", "#5ab4ac"),
                    labels = c('Clean election', 'Freedom of association', 'Freedom of expression')) + 
  theme_minimal() +
  theme(plot.caption = element_text(hjust = 0))
  
```

### United States

```{r}
vdemdata_long %>%
  filter(country_name == 'United States of America') %>% 
  ggplot(aes(x = year, y = score, color = criteria)) +
  geom_point() +
  geom_line() + 
  labs(title = 'Electoral Democracy in the USA from 1971 to 2022',
       subtitle = 'Trends in electoral democracy index scores over the last 50 years',
       x = 'Year',
       y = 'Index Score',
       color = 'Index',
       caption = 'Data Source: Varieties of Democracy (https://www.v-dem.net)') +
  scale_color_manual(values = c("#d8b365", "#2c7fb8", "#5ab4ac"),
                    labels = c('Clean election', 'Freedom of association', 'Freedom of expression')) + 
  theme_minimal() +
  theme(plot.caption = element_text(hjust = 0))
  
```

## Cross-Country Results

### Individual measures of electoral democracy

```{r}

#| Summary of individual measures of electoral democracy

indicators = c('clean_elections' = 'Clean election', 
               'freedom_association' = 'Freedom of association', 
               'freedom_expression' = 'Freedom of expression')

vdemdata_long %>%
  ggplot(aes(y = country_name, x = score)) +
  geom_bar(stat = "summary", fill = '#008080') +
  facet_wrap(~criteria, labeller = labeller(criteria = indicators)) +
  labs(title = 'Electoral democracy indicators in Bangladesh, Pakistan, Philippines, and USA', 
       subtitle = 'Comparison of individual index scores',
       x = 'Index Score',
       y = 'Country',
       caption = 'Data Source: Varieties of Democracy (https://www.v-dem.net)') +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0),
        plot.caption = element_text(hjust = 0))

```

### Aggregated measure of electoral democracy

```{r}
# calculate index score
vdemdata_scored <- vdemdata_long %>%
  pivot_wider( # convert the data to a wide format to convert all criteria back as columns 
    names_from = "criteria",
    values_from = "score",
  ) %>%
  mutate(electoral_democracy = (0.5 * # calculate based on VDEM formula of electoral democracy index
                                  (elected_index *
                                     clean_elections *
                                     freedom_association *
                                     freedom_expression *
                                     suffrage)) +
           (0.5 * ((.125 * elected_index) +
                     (.25*clean_elections) +
                     (.25 * freedom_association) +
                     (.125 * suffrage) +
                     (.25 * freedom_expression))))
```

```{r}
#| Descriptive statistics

vdemdata_scored %>%
  group_by(country_name) %>%
  summarise(yearly_average = mean(electoral_democracy),
            yearly_median = median(electoral_democracy),
            sd = sd(electoral_democracy),
            min = min(electoral_democracy),
            max = min(electoral_democracy))



```

```{r}
#| Aggregate score visualization

vdemdata_scored %>%
  ggplot(aes(x = year, y = electoral_democracy, color = country_name)) +
  geom_point() +
  geom_line() +
  theme_minimal()

```

# Comparing three countries' mean democracy index scores with that of the United States using two-sample t-test

### Bangladesh

```{r}
us_vdemdata <- vdemdata_scored %>%
  filter(country_name == 'United States of America')

bd_vdemdata <- vdemdata_scored %>%
  filter(country_name == 'Bangladesh')

ttest_bd <- t.test(us_vdemdata$electoral_democracy,bd_vdemdata$electoral_democracy)
print(ttest_bd)

```

For this two-sample t-test, we consider

**null hypothesis, Ho:** there is no difference in the electoral democracy index scores between Bangladesh and the United States. **alternative hypothesis, Ha:** there is a difference between the electoral democracy index scores between these two countries.

While advancing to our t-test, we find an extremely **low p-value** *(\<2.2e-16)* which gives us strong evidence to reject the null hypothesis. Therefore, we conclude that there is a statistically significant difference in the mean electoral democracy index scores between Bangladesh and the United States.

### Pakistan

```{r}

pk_vdemdata <- vdemdata_scored %>%
  filter(country_name == 'Pakistan')

ttest_pk <- t.test(us_vdemdata$electoral_democracy,pk_vdemdata$electoral_democracy)
print(ttest_pk)

```

For this two-sample t-test, we consider

**null hypothesis, Ho:** there is no difference in the electoral democracy index scores between Pakistan and the United States. **alternative hypothesis, Ha:** there is a difference between the electoral democracy index scores between these two countries.

While advancing to our t-test, we find an extremely **low p-value** *(\<2.2e-16)* which gives us strong evidence to reject the null hypothesis. Therefore, we conclude that there is a statistically significant difference in the mean electoral democracy index scores between Pakistan and the United States.

### Philippines

```{r}
ph_vdemdata <- vdemdata_scored %>%
  filter(country_name == 'Philippines')

ttest_ph <- t.test(us_vdemdata$electoral_democracy,ph_vdemdata$electoral_democracy)
print(ttest_ph)

```

For this two-sample t-test, we consider

**null hypothesis, Ho:** there is no difference in the electoral democracy index scores between Philippines and the United States. **alternative hypothesis, Ha:** there is a difference between the electoral democracy index scores between these two countries.

While advancing to our t-test, we find an extremely **low p-value** *(\<2.2e-16)* which gives us strong evidence to reject the null hypothesis. Therefore, we conclude that there is a statistically significant difference in the mean electoral democracy index scores between Philippines and the United States.

# Discussion and Conclusion

# References
