---
title: "EDLD 651 Final Project"
author: "Jon Benedik Bunquin, Nafisul Huq, and Faria Shaikh"
format: html
editor: visual
warning: False
embed-resources: true
toc: true
bibliography: references.bib
---

# Abstract

# Introduction

**Background**

The geopolitical landscape following post-World War II era witnessed a huge shift with the global wave of decolonization. This change led to to the birth of many newly sovereign states. These new countries, or "*young democracies*" embarked on the path of democratic governance for the very first time.

According to @barlas2019 these nascent democracies, emerging from the shadows of colonial rule are facing one big challenge: to establish structured, and stable democratic systems in their countries.

Moreover, as shared by @lin2020, these democracies, in their formative stages, and even today, are encountering significant challenges, which includes institutional fragility , military coups, corrupt power systems, bad governance, and vulnerability to local and foreign pressures. These pressures led to significant concerns about "democratic backsliding" - a term that was coined particularly to study the regressive shift from democratic norms while create a way for authoritarian regimes.

**Review of Literature**

**a. Decolonization and Democracy**

@smith2021 offers an detailed overview of political transformations following decolonization, while emphasizing the complexities faced by new states in establishing democratic governance.

He suggests that the biggest challenges arise from a lack of ethnic diversity, economic challenges, and the legacies of colonial rule when establishing a democratic regime. While not directly focused on decolonization, @datta2018's work contributes to understanding the broader context of political transformations and democratization in post-colonial states. His work also offers insights into the historical context and processes of decolonization, providing a foundation of the subject and it's impact on subsequent democracies.

Moreover, the enduring influence of colonization on institutional frameworks and governance structures is emphasized by @spiegeletal2016 . The study also shows that the existence of representative political bodies during the late colonial period clearly predicts the existence and stability of democracy in recent decades, which also highlights the long-term impact of colonial governance on democratic institutions in developing countries.

While not directly related to democratic government, @sultana2019 explores the broader implications of decolonization that goes beyond political ideologies and systems, and have a deeper and profound impact on societies, including culture, values and classes.

**b. Democratic Backsliding**

The concept of democratic backsliding has gained a lot of attention recently as it focuses on understanding the erosion of democratic norms and institutions. The term "democratic backsliding" is used to describe a progressive deterioration in the quality of democracy. This phenomenon occurs when essential democratic institutions become weakened or are exploited for purposes contrary to their intended democratic function. Such degradation in democratic processes and structures often transpires incrementally and may not be immediately apparent, akin to structural weaknesses developing in a building over time.

@bermeo2016 discusses democratic backsliding as it leads to gradual decline in the quality of democracy. This happens when key institutions meant to support democracy become weak or are misused. In particular, the phenomenon of democratic backsliding in developing countries has gained significant attention from political scientists, and developing country scholars. The focus has been to understand the processes, implications, and potential mitigating factors associated with it.

@luhrmann2019 critically studied democratic backsliding, and highlighted the challenges and nuances associated with understanding the complexities that are involved in assessing democratic decline in developing countries.

Apart from exploring the impact of decolonization on democracy in an entire region, it is also very important to study democratic backsliding from perspective of independent countries. @ruud2018 delves into Bangladesh's political and democratic journey, focusing on its struggles with military rule and democratic governance. This country-specific analysis provides insights into the unique challenges and trajectories of Bangladesh's democracy, contributing to the understanding of democratic backsliding in the context of Bangladesh. While @jalal2014 offers a critical analysis of Pakistan's democratic evolution, highlighting the interplay between civilian governance and military power. Similarly, @hutchcroft2016 examines the role of political dynasties and the influence of populist leadership in shaping the Philippines' democratic framework.

**Comparative Analysis of Democracies**

@lijphart2013's work "Patterns of Democracy" provides a comprehensive framework for understanding different models of democracies and their comparative performances. In particular, there are individual and very nuanced factors that impact the current state of democracy in each one of these countries. From the context of Bangladesh, @khan2014 analyze the the compatibility of democracy and good governance in Bangladesh, and emphasize on the need for a stable democratic regime in the country while exploring the reasons it has not been successful so far. Moreover, @riaz2021 also highlights the reasons behind the separation of Pakistan and Bangladesh while shedding light on how it has caused democratic backsliding in both the countries.

Furthermore, when it comes to Pakistan, @khalid2023 discusses the present situation and the current dilemma of democracy in Pakistan; while focusing on the reasons behind failure of democracy in the relatively young country despite many attempts at establishing a successful democratic government.

@hoffman2011 also talks about the complex dynamics of military extrication and temporary democracy in Pakistan. He also focuses on the relationship between military intervention and democratic governance.

Finally, when it comes to Philippines, @hutchcroft2003 studied the origins and evolution of the democratic foundation in the country while offering a detailed analysis of the various events that impacted the history of democracy in the country. @lorch2021 also studies why democracies get weaker,and what role do the civil society organizations (CSOs) play. His study from the context of developing countries show that CSOs do not have a major impact on the success of democracy, complicating it even further.

To conclude, reviewing the literature on stability of democracy in developing countries including Bangladesh, Pakistan, and the Philippines help in gaining a deeper understanding of the challenges, dynamics and future prospects of the democratic governance in these countries.

# Methods

```{r}
## Imports

#| echo: false

#install.packages("devtools")
#install.packages("vdemdata")

library(vdemdata)
library(dplyr)
library(skimr)
library(tidyverse)
library(ggplot2)
library(stats)
vdem <- vdem
```

## Data Processing

### Data Filtering

```{r}
vdemdata <- vdem %>%
  filter(country_name == "Bangladesh" | 
           country_name == "Philippines" |
           country_name == "Pakistan" |
           country_name == "United States of America") %>%
  select(country_name, 
         year, 
         v2x_frassoc_thick,
         v2xel_frefair,
         v2x_freexp_altinf,
         v2x_elecoff,
         v2x_suffr) %>%
  filter(year >= 1971) %>% 
  rename(freedom_association=v2x_frassoc_thick,
         clean_elections=v2xel_frefair,
         freedom_expression=v2x_freexp_altinf,
         elected_index=v2x_elecoff,
         suffrage=v2x_suffr)

head(vdemdata)

```

### Data Exploration

```{r}
skim(vdemdata)
```

<insert interpretation here>

### Data Cleaning

```{r}
vdemdata_long <- vdemdata %>%
  pivot_longer(
    cols = 3:5,
    names_to = 'criteria',
    values_to = 'score'
    )

head(vdemdata_long)
```

# Results

## Individual Country Results

### Bangladesh

```{r}
vdemdata_long %>%
  filter(country_name == 'Bangladesh') %>% 
  ggplot(aes(x = year, y = score, color = criteria)) +
  geom_point() +
  geom_line() + 
  geom_vline(xintercept = c(1975,1980,1990,2000,2007,2013)) + 
  annotate("text", x=c(1974,1979,1989,1999,2006,2012), y=0.1,label=c("Before 1975","1980","1990","2000","2007","2013"), 
           color="tomato4",size=2.5, angle=90) + 
  labs(title = 'Electoral Democracy in the Bangladesh from 1971 to 2022',
       subtitle = 'Trends in electoral democracy index scores over the last 50 years',
       x = 'Year',
       y = 'Index Score',
       color = 'Index',
       caption = 'Data Source: Varieties of Democracy (https://www.v-dem.net)') +
  scale_color_manual(values = c("#d8b365", "#2c7fb8", "#5ab4ac"),
                    labels = c('Clean election', 'Freedom of association', 'Freedom of expression')) + 
  theme_minimal() +
  theme(plot.caption = element_text(hjust = 0))
  
```

The above graph of Bangladesh shows an overview of the country's electoral democracy status from **1970 to 2022** focusing on three key variables - freedom of association, freedom of expression, and clean election. It reveals an unwavering trend of electoral democracy in the country ultimately leaning towards a declining state with a cumulative effect of the variables, where freedom of association is seen in a consistent battle to stay resilient in its consistent unstable trajectory, freedom of expression is found experiencing the most fluctuations with more decline and less rise in scores throughout the interval, and clean election is seen affected by the most severe slumps.\
</br> The ups and downs of the variables include many significant events that impacted the democracy in Bangladesh in the given period. To begin with, Bangladesh witnessed the first blow in **1975** with the formation of a one-party government that threatened the country's free and fair election, freedom of expression, and freedom of association. Then, it saw the emergence of military rule in late **1975** against the backdrop of tyrannical rule lasting through **1980**. Although sounds weird, it appeared as a pathway to get back to democracy. After years of smooth functioning, the military rule transformed into a figure of hatred and disdain. As a result, Bangladesh saw the fall of military dictatorship and the rise of democracy from **1990 to 2000**. The next bar towards the country's democracy became apparent in **2007** when an interim government came into play to offset that bar. While democracy sustained after that period it did not sustain for long. Greed of the erstwhile ruling party to stay in power battered the foundations of democracy, including clean election, freedom of speech, and freedom of association along with other components. The consequence of this event was so adverse that the country kept struggling to ascend the ladder of democracy until recently.

### Pakistan

```{r}

vdemdata_long %>%
  filter(country_name == 'Pakistan') %>% 
  ggplot(aes(x = year, y = score, color = criteria)) +
  geom_point() +
  geom_line() + geom_vline(xintercept = c(1971,1977,1988,1999,2008)) + 
  annotate("text", x=c(1970,1976,1986,1998,2007), y=0.1,label=c("Before 1971","1977","1988","1999","2008"),
           color="tomato4",size=2.5, angle=90) +
  labs(title = 'Electoral Democracy in the Pakistan from 1971 to 2022',
       subtitle = 'Trends in electoral democracy index scores over the last 50 years',
       x = 'Year',
       y = 'Index Score',
       color = 'Index',
       caption = 'Data Source: Varieties of Democracy (https://www.v-dem.net)') +
  scale_color_manual(values = c("#d8b365", "#2c7fb8", "#5ab4ac"),
                    labels = c('Clean election', 'Freedom of association', 'Freedom of expression')) + 
  theme_minimal() +
  theme(plot.caption = element_text(hjust = 0))

```

### Philippines

```{r}
vdemdata_long %>%
  filter(country_name == 'Philippines') %>% 
  ggplot(aes(x = year, y = score, color = criteria)) +
  geom_point() +
  
  geom_line() + 
  geom_vline(xintercept = c(1970,1986,2004,2009,2016)) + 
  annotate("text", x=c(1969,1985,2003,2008,2015), y=0.1,label=c("1970","1986","2004","2009","2016"), 
           color="tomato4",size=2.5, angle=90) +
  labs(title = 'Electoral Democracy in the Philippines from 1971 to 2022',
       subtitle = 'Trends in electoral democracy index scores over the last 50 years',
       x = 'Year',
       y = 'Index Score',
       color = 'Index',
       caption = 'Data Source: Varieties of Democracy (https://www.v-dem.net)') +
  scale_color_manual(values = c("#d8b365", "#2c7fb8", "#5ab4ac"),
                    labels = c('Clean election', 'Freedom of association', 'Freedom of expression')) + 
  theme_minimal() +
  theme(plot.caption = element_text(hjust = 0))
  
```

### United States

```{r}
vdemdata_long %>%
  filter(country_name == 'United States of America') %>% 
  ggplot(aes(x = year, y = score, color = criteria)) +
  geom_point() +
  geom_line() + 
  labs(title = 'Electoral Democracy in the USA from 1971 to 2022',
       subtitle = 'Trends in electoral democracy index scores over the last 50 years',
       x = 'Year',
       y = 'Index Score',
       color = 'Index',
       caption = 'Data Source: Varieties of Democracy (https://www.v-dem.net)') +
  scale_color_manual(values = c("#d8b365", "#2c7fb8", "#5ab4ac"),
                    labels = c('Clean election', 'Freedom of association', 'Freedom of expression')) + 
  theme_minimal() +
  theme(plot.caption = element_text(hjust = 0))
  
```

## Cross-Country Results

### Individual measures of electoral democracy

```{r}

#| Summary of individual measures of electoral democracy

indicators = c('clean_elections' = 'Clean election', 
               'freedom_association' = 'Freedom of association', 
               'freedom_expression' = 'Freedom of expression')

vdemdata_long %>%
  ggplot(aes(y = country_name, x = score)) +
  geom_bar(stat = "summary", fill = '#008080') +
  facet_wrap(~criteria, labeller = labeller(criteria = indicators)) +
  labs(title = 'Electoral democracy indicators in Bangladesh, Pakistan, Philippines, and USA', 
       subtitle = 'Comparison of individual index scores',
       x = 'Index Score',
       y = 'Country',
       caption = 'Data Source: Varieties of Democracy (https://www.v-dem.net)') +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0),
        plot.caption = element_text(hjust = 0))

```

### Aggregated measure of electoral democracy

```{r}
# calculate index score
vdemdata_scored <- vdemdata_long %>%
  pivot_wider( # convert the data to a wide format to convert all criteria back as columns 
    names_from = "criteria",
    values_from = "score",
  ) %>%
  mutate(electoral_democracy = (0.5 * # calculate based on VDEM formula of electoral democracy index
                                  (elected_index *
                                     clean_elections *
                                     freedom_association *
                                     freedom_expression *
                                     suffrage)) +
           (0.5 * ((.125 * elected_index) +
                     (.25*clean_elections) +
                     (.25 * freedom_association) +
                     (.125 * suffrage) +
                     (.25 * freedom_expression))))
```

```{r}
#| Descriptive statistics

vdemdata_scored %>%
  group_by(country_name) %>%
  summarise(yearly_average = mean(electoral_democracy),
            yearly_median = median(electoral_democracy),
            sd = sd(electoral_democracy),
            min = min(electoral_democracy),
            max = min(electoral_democracy))



```

```{r}
#| Aggregate score visualization

vdemdata_scored %>%
  ggplot(aes(x = year, y = electoral_democracy, color = country_name)) +
  geom_point() +
  geom_line() +
  theme_minimal()

```

# Comparing three countries' mean democracy index scores with that of the United States using two-sample t-test

### Bangladesh

```{r}
us_vdemdata <- vdemdata_scored %>%
  filter(country_name == 'United States of America')

bd_vdemdata <- vdemdata_scored %>%
  filter(country_name == 'Bangladesh')

ttest_bd <- t.test(us_vdemdata$electoral_democracy,bd_vdemdata$electoral_democracy)
print(ttest_bd)

```

For this two-sample t-test, we consider

**null hypothesis, Ho:** there is no difference in the electoral democracy index scores between Bangladesh and the United States. **alternative hypothesis, Ha:** there is a difference between the electoral democracy index scores between these two countries.

While advancing to our t-test, we find an extremely **low p-value** *(\<2.2e-16)* which gives us strong evidence to reject the null hypothesis. Therefore, we conclude that there is a statistically significant difference in the mean electoral democracy index scores between Bangladesh and the United States.

### Pakistan

```{r}

pk_vdemdata <- vdemdata_scored %>%
  filter(country_name == 'Pakistan')

ttest_pk <- t.test(us_vdemdata$electoral_democracy,pk_vdemdata$electoral_democracy)
print(ttest_pk)

```

For this two-sample t-test, we consider

**null hypothesis, Ho:** there is no difference in the electoral democracy index scores between Pakistan and the United States. **alternative hypothesis, Ha:** there is a difference between the electoral democracy index scores between these two countries.

While advancing to our t-test, we find an extremely **low p-value** *(\<2.2e-16)* which gives us strong evidence to reject the null hypothesis. Therefore, we conclude that there is a statistically significant difference in the mean electoral democracy index scores between Pakistan and the United States.

### Philippines

```{r}
ph_vdemdata <- vdemdata_scored %>%
  filter(country_name == 'Philippines')

ttest_ph <- t.test(us_vdemdata$electoral_democracy,ph_vdemdata$electoral_democracy)
print(ttest_ph)

```

For this two-sample t-test, we consider

**null hypothesis, Ho:** there is no difference in the electoral democracy index scores between Philippines and the United States. **alternative hypothesis, Ha:** there is a difference between the electoral democracy index scores between these two countries.

While advancing to our t-test, we find an extremely **low p-value** *(\<2.2e-16)* which gives us strong evidence to reject the null hypothesis. Therefore, we conclude that there is a statistically significant difference in the mean electoral democracy index scores between Philippines and the United States.

# Discussion and Conclusion

# References
