---
title: "EDLD 651 Final Project"
author: "Jon Benedik Bunquin, Nafisul Huq, and Faria Shaikh"
format: html
editor: visual
warning: False
embed-resources: true
toc: true
bibliography: references.bib
---

# Abstract

# Introduction

**Introduction**

The geopolitical landscape of the post-World War II era witnessed a monumental shift with the global wave of decolonization, especially pronounced in the 1960s. This period marked the birth of numerous newly sovereign states, many of which embarked on the path of democratic governance. These nascent democracies, emerging from the shadows of
colonial rule, faced the daunting task of establishing democratic systems, cultivating a political culture rooted in democratic principles, and nurturing values conducive to a democratic ethos. However, as illuminated by Kapstein and Converse (2008), these democracies, in their formative stages, encountered
significant challenges, including institutional fragility and vulnerability to internal and external pressures, leading to concerns about democratic backsliding -- a regressive shift from democratic norms that could potentially pave the way for authoritarian regimes.

**Review of Literature**

The study of electoral democracy over the past fifty years, especially in the context of post-colonial countries and comparative analyses with established democracies like the United States,
is a rich field of inquiry. This literature review will explore key themes including the global context of democratic evolution post-decolonization, the specific democratic trajectories of Bangladesh, Pakistan, and the Philippines, methodological approaches in democracy research, and predictive models in
political science.

# Methods

```{r}
## Imports

#| echo: false

#install.packages("devtools")
#install.packages("vdemdata")

library(vdemdata)
library(dplyr)
library(skimr)
library(tidyverse)
library(ggplot2)
library(stats)
vdem <- vdem
```

## Data Processing

### Data Filtering

```{r}
vdemdata <- vdem %>%
  filter(country_name == "Bangladesh" | 
           country_name == "Philippines" |
           country_name == "Pakistan" |
           country_name == "United States of America") %>%
  select(country_name, 
         year, 
         v2x_frassoc_thick,
         v2xel_frefair,
         v2x_freexp_altinf,
         v2x_elecoff,
         v2x_suffr) %>%
  filter(year >= 1971) %>% 
  rename(freedom_association=v2x_frassoc_thick,
         clean_elections=v2xel_frefair,
         freedom_expression=v2x_freexp_altinf,
         elected_index=v2x_elecoff,
         suffrage=v2x_suffr)

head(vdemdata)

```

### Data Exploration

```{r}
skim(vdemdata)
```

<insert interpretation here>

### Data Cleaning

```{r}
vdemdata_long <- vdemdata %>%
  pivot_longer(
    cols = 3:5,
    names_to = 'criteria',
    values_to = 'score'
    )

head(vdemdata_long)
```

# Results

## Individual Country Results

### Bangladesh

```{r}
vdemdata_long %>%
  filter(country_name == 'Bangladesh') %>% 
  ggplot(aes(x = year, y = score, color = criteria)) +
  geom_point() +
  geom_line() + 
  geom_vline(xintercept = c(1975,1980,1990,2000,2007,2013)) + 
  annotate("text", x=c(1974,1979,1989,1999,2006,2012), y=0.1,label=c("Before 1975","1980","1990","2000","2007","2013"), 
           color="tomato4",size=2.5, angle=90) + 
  labs(title = 'Electoral Democracy in the Bangladesh from 1971 to 2022',
       subtitle = 'Trends in electoral democracy index scores over the last 50 years',
       x = 'Year',
       y = 'Index Score',
       color = 'Index',
       caption = 'Data Source: Varieties of Democracy (https://www.v-dem.net)') +
  scale_color_manual(values = c("#d8b365", "#2c7fb8", "#5ab4ac"),
                    labels = c('Clean election', 'Freedom of association', 'Freedom of expression')) + 
  theme_minimal() +
  theme(plot.caption = element_text(hjust = 0))
  
```

The above graph of Bangladesh shows an overview of the country's electoral democracy status from **1970 to 2022** focusing on three key variables - freedom of association, freedom of expression, and clean election. It reveals an unwavering trend of electoral democracy in the country ultimately leaning towards a declining state with a cumulative effect of the variables, where freedom of association is seen in a consistent battle to stay resilient in its consistent unstable trajectory, freedom of expression is found experiencing the most fluctuations with more decline and less rise in scores throughout the interval, and clean election is seen affected by the most severe slumps.\
</br> The ups and downs of the variables include many significant events that impacted the democracy in Bangladesh in the given period. To begin with, Bangladesh witnessed the first blow in **1975** with the formation of a one-party government that threatened the country's free and fair election, freedom of expression, and freedom of association. Then, it saw the emergence of military rule in late **1975** against the backdrop of tyrannical rule lasting through **1980**. Although sounds weird, it appeared as a pathway to get back to democracy. After years of smooth functioning, the military rule transformed into a figure of hatred and disdain. As a result, Bangladesh saw the fall of military dictatorship and the rise of democracy from **1990 to 2000**. The next bar towards the country's democracy became apparent in **2007** when an interim government came into play to offset that bar. While democracy sustained after that period it did not sustain for long. Greed of the erstwhile ruling party to stay in power battered the foundations of democracy, including clean election, freedom of speech, and freedom of association along with other components. The consequence of this event was so adverse that the country kept struggling to ascend the ladder of democracy until recently.

### Pakistan

```{r}

vdemdata_long %>%
  filter(country_name == 'Pakistan') %>% 
  ggplot(aes(x = year, y = score, color = criteria)) +
  geom_point() +
  geom_line() + geom_vline(xintercept = c(1971,1977,1988,1999,2008)) + 
  annotate("text", x=c(1970,1976,1986,1998,2007), y=0.1,label=c("Before 1971","1977","1988","1999","2008"),
           color="tomato4",size=2.5, angle=90) +
  labs(title = 'Electoral Democracy in the Pakistan from 1971 to 2022',
       subtitle = 'Trends in electoral democracy index scores over the last 50 years',
       x = 'Year',
       y = 'Index Score',
       color = 'Index',
       caption = 'Data Source: Varieties of Democracy (https://www.v-dem.net)') +
  scale_color_manual(values = c("#d8b365", "#2c7fb8", "#5ab4ac"),
                    labels = c('Clean election', 'Freedom of association', 'Freedom of expression')) + 
  theme_minimal() +
  theme(plot.caption = element_text(hjust = 0))

```

### Philippines

```{r}
vdemdata_long %>%
  filter(country_name == 'Philippines') %>% 
  ggplot(aes(x = year, y = score, color = criteria)) +
  geom_point() +
  
  geom_line() + 
  geom_vline(xintercept = c(1970,1986,2004,2009,2016)) + 
  annotate("text", x=c(1969,1985,2003,2008,2015), y=0.1,label=c("1970","1986","2004","2009","2016"), 
           color="tomato4",size=2.5, angle=90) +
  labs(title = 'Electoral Democracy in the Philippines from 1971 to 2022',
       subtitle = 'Trends in electoral democracy index scores over the last 50 years',
       x = 'Year',
       y = 'Index Score',
       color = 'Index',
       caption = 'Data Source: Varieties of Democracy (https://www.v-dem.net)') +
  scale_color_manual(values = c("#d8b365", "#2c7fb8", "#5ab4ac"),
                    labels = c('Clean election', 'Freedom of association', 'Freedom of expression')) + 
  theme_minimal() +
  theme(plot.caption = element_text(hjust = 0))
  
```

### United States

```{r}
vdemdata_long %>%
  filter(country_name == 'United States of America') %>% 
  ggplot(aes(x = year, y = score, color = criteria)) +
  geom_point() +
  geom_line() + 
  labs(title = 'Electoral Democracy in the USA from 1971 to 2022',
       subtitle = 'Trends in electoral democracy index scores over the last 50 years',
       x = 'Year',
       y = 'Index Score',
       color = 'Index',
       caption = 'Data Source: Varieties of Democracy (https://www.v-dem.net)') +
  scale_color_manual(values = c("#d8b365", "#2c7fb8", "#5ab4ac"),
                    labels = c('Clean election', 'Freedom of association', 'Freedom of expression')) + 
  theme_minimal() +
  theme(plot.caption = element_text(hjust = 0))
  
```

## Cross-Country Results

### Individual measures of electoral democracy

```{r}

#| Summary of individual measures of electoral democracy

indicators = c('clean_elections' = 'Clean election', 
               'freedom_association' = 'Freedom of association', 
               'freedom_expression' = 'Freedom of expression')

vdemdata_long %>%
  ggplot(aes(y = country_name, x = score)) +
  geom_bar(stat = "summary", fill = '#008080') +
  facet_wrap(~criteria, labeller = labeller(criteria = indicators)) +
  labs(title = 'Electoral democracy indicators in Bangladesh, Pakistan, Philippines, and USA', 
       subtitle = 'Comparison of individual index scores',
       x = 'Index Score',
       y = 'Country',
       caption = 'Data Source: Varieties of Democracy (https://www.v-dem.net)') +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0),
        plot.caption = element_text(hjust = 0))

```

### Aggregated measure of electoral democracy

```{r}
# calculate index score
vdemdata_scored <- vdemdata_long %>%
  pivot_wider( # convert the data to a wide format to convert all criteria back as columns 
    names_from = "criteria",
    values_from = "score",
  ) %>%
  mutate(electoral_democracy = (0.5 * # calculate based on VDEM formula of electoral democracy index
                                  (elected_index *
                                     clean_elections *
                                     freedom_association *
                                     freedom_expression *
                                     suffrage)) +
           (0.5 * ((.125 * elected_index) +
                     (.25*clean_elections) +
                     (.25 * freedom_association) +
                     (.125 * suffrage) +
                     (.25 * freedom_expression))))
```

```{r}
#| Descriptive statistics

vdemdata_scored %>%
  group_by(country_name) %>%
  summarise(yearly_average = mean(electoral_democracy),
            yearly_median = median(electoral_democracy),
            sd = sd(electoral_democracy),
            min = min(electoral_democracy),
            max = min(electoral_democracy))



```

```{r}
#| Aggregate score visualization

vdemdata_scored %>%
  ggplot(aes(x = year, y = electoral_democracy, color = country_name)) +
  geom_point() +
  geom_line() +
  theme_minimal()

```

# Comparing three countries' mean democracy index scores with that of the United States using two-sample t-test

### Bangladesh

```{r}
us_vdemdata <- vdemdata_scored %>%
  filter(country_name == 'United States of America')

bd_vdemdata <- vdemdata_scored %>%
  filter(country_name == 'Bangladesh')

ttest_bd <- t.test(us_vdemdata$electoral_democracy,bd_vdemdata$electoral_democracy)
print(ttest_bd)

```

For this two-sample t-test, we consider

**null hypothesis, Ho:** there is no difference in the electoral democracy index scores between Bangladesh and the United States. **alternative hypothesis, Ha:** there is a difference between the electoral democracy index scores between these two countries.

While advancing to our t-test, we find an extremely **low p-value** *(\<2.2e-16)* which gives us strong evidence to reject the null hypothesis. Therefore, we conclude that there is a statistically significant difference in the mean electoral democracy index scores between Bangladesh and the United States.

### Pakistan

```{r}

pk_vdemdata <- vdemdata_scored %>%
  filter(country_name == 'Pakistan')

ttest_pk <- t.test(us_vdemdata$electoral_democracy,pk_vdemdata$electoral_democracy)
print(ttest_pk)

```

For this two-sample t-test, we consider

**null hypothesis, Ho:** there is no difference in the electoral democracy index scores between Pakistan and the United States. **alternative hypothesis, Ha:** there is a difference between the electoral democracy index scores between these two countries.

While advancing to our t-test, we find an extremely **low p-value** *(\<2.2e-16)* which gives us strong evidence to reject the null hypothesis. Therefore, we conclude that there is a statistically significant difference in the mean electoral democracy index scores between Pakistan and the United States.

### Philippines

```{r}
ph_vdemdata <- vdemdata_scored %>%
  filter(country_name == 'Philippines')

ttest_ph <- t.test(us_vdemdata$electoral_democracy,ph_vdemdata$electoral_democracy)
print(ttest_ph)

```

For this two-sample t-test, we consider

**null hypothesis, Ho:** there is no difference in the electoral democracy index scores between Philippines and the United States. **alternative hypothesis, Ha:** there is a difference between the electoral democracy index scores between these two countries.

While advancing to our t-test, we find an extremely **low p-value** *(\<2.2e-16)* which gives us strong evidence to reject the null hypothesis. Therefore, we conclude that there is a statistically significant difference in the mean electoral democracy index scores between Philippines and the United States.

# Discussion and Conclusion

# References
